---
title: "Welcome to the Class"
author: "Ian McCarthy | Emory University"
format: 
  revealjs:
    theme: [moon]
    preview-links: auto
    chalkboard:
      boardmarker-width: 5
    slide-number: true
    width: 1600
    height: 900    
#    embed-resources: true
from: markdown+emoji
execute: 
  echo: true
---

```{r}
#| include: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, lubridate, gapminder, scales,
               hrbrthemes, gganimate, OECD, here)
```

## Motivating question

**Does health care spending improve health?**

Let's motivate the course with this simple question...does health care spending improve our health?

---

## Spending and Health

```{r}
#| include: false
source(here("00-introduction/DartmouthAtlas.R"))
```


```{r}
#| code-fold: true
#| code-summary: "R Code"
#| fig-align: center
ggplot(data = (dartmouth.data %>% filter(Year==2015)), 
       mapping = aes(x = Expenditures, y = Total_Mortality)) + 
  geom_point(size = 1) + theme_bw() + scale_x_continuous(label = comma) +
  geom_smooth(method="lm", se=FALSE, color="blue", size=1/2) +
  labs(x = "Spending Per Capita ($US)",
       y = "Mortality Rate",
       title = "Mortality and Health Care Spending")
```


---

## Spending and Health

- Does medical spending make us sicker?
- What else might explain this relationship?


---

## Goals of this course.

1. Understand and implement selected methods for causal inference
2. Along the way...data management and version control with real data
3. Summarize, visualize, and explain research results


# Syllabus highlights


---

## Why this course?

1. Major problems that need solutions
2. Need good, convincing empirical work for policy
3. Working with data is hard, particularly health care data
4. Your work should be transparent and reproducible

---

## Structure

- Very applied in nature
- *Methods* for causal inference
  - Selection on observables (regression, re-weighting, matching, propensity scores)
  - Instrumental variables
  - Regression discontinuity
  - Difference-in-differences


---

## Structure

- *Substantive* areas
  - Hospital pricing, policy, and competition
  - Cigarette taxes and demand
  - Medicare Advantage and quality disclosure
  - Medicaid expansion and health insurance


---

## Structure

- *Datasets* from the real world
  - Hospital Cost Report Information System (HCRIS)
  - Centers for Disease Control (CDC)
  - Medicare Advantage data
  - Behavioral Risk Factor Surveillance System (BRFSS), Medicaid, Health Insurance Exchanges
  
  
---

## Assignments

- Homework (x5)
- Peer Reviews (x4)
- Participation

---

## Grading

| Component                           | Weight |
|:-|-:|
| 5 × homework assignments (15% each) | 75% |
| Peer review                   	  | 20% |
| Participation  					  | 5%  |
